---
title: "exercise-08"
format: html
editor: visual
---
Step 1: Load and explore data
```{r}
library(tidyverse)
library(dplyr)
library(skimr)
f<- "https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/Street_et_al_2017.csv" #loads in my file
d<- read_csv(f, col_names = TRUE) #reads in my dataset 
glimpse(d) #brief look at the data
skim(d) #gives me summary statistics
```

Step 2:From this dataset, plot brain size (ECV) as a function of social group size (Group_size), longevity (Longevity), juvenile period length (Weaning), and reproductive lifespan (Repro_lifespan).
```{r}
library(ggplot2)
library(cowplot) #for plot_grid()

# Group Size
p1<-ggplot(d, aes(x = Group_size, y = ECV)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "ECV vs Group Size")

# Longevity
p2<-ggplot(d, aes(x = Longevity, y = ECV)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "ECV vs Longevity")

# Weaning
p3<- ggplot(d, aes(x = Weaning, y = ECV)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "ECV vs Weaning")

# Reproductive Lifespan
p4<-ggplot(d, aes(x = Repro_lifespan, y = ECV)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "ECV vs Reproductive Lifespan")

p5<-plot_grid(p1,p2,p3,p4, nrow = 2) #plots them together
```

Step 3: Derive by hand the ordinary least squares regression coefficients for ECV as a function of social group size.
```{r}
library(tidyr)

d1 <- d %>%
  drop_na(ECV, Group_size) #removes NAs 

#define variables
x <- d1$Group_size
y <- d1$ECV

#OLS equation
x_mean <- mean(x)
y_mean <- mean(y)

beta1 <- sum((x - x_mean) * (y - y_mean)) / sum((x - x_mean)^2)
beta0 <- y_mean - beta1 * x_mean

beta0
beta1
```

Step 4: Confirm that you get the same results using the lm() function.
```{r}
lm(ECV ~ Group_size, data = d1) #check results are the same
```
Results are the same!

Step 5: Repeat the analysis above for three different major radiations of primates - “catarrhines”, “platyrrhines”, and “strepsirhines”) separately. These are stored in the variable Taxonomic_group. Do your regression coefficients differ among groups? How might you determine this?
```{r}
d2 <- d %>%
  drop_na(ECV, Group_size, Taxonomic_group)

model_cat <- lm(ECV ~ Group_size,
                data = filter(d2, Taxonomic_group == "Catarrhini"))

model_plat <- lm(ECV ~ Group_size,
                 data = filter(d2, Taxonomic_group == "Platyrrhini"))

model_strep <- lm(ECV ~ Group_size,
                  data = filter(d2, Taxonomic_group == "Strepsirhini"))

summary(model_cat)
summary(model_plat)
summary(model_strep)

#checking for statistacal differences in slope
model_interaction <- lm(ECV ~ Group_size * Taxonomic_group, data = d2)
summary(model_interaction)
```
Regression coefficients differed among primate radiations. Catarrhines showed the weakest relationship between group size and brain size (β₁ = 1.146), while platyrrhines and strepsirhines showed stronger, more similar slopes (β₁ ≈ 1.96 and 1.84, respectively). These differences can be further tested using a linear model with an interaction term (ECV ~ Group_size × Taxonomic_group), where significant interaction terms indicate that slopes differ among groups. Once completed, we see that while they do initially appear different, they are not statistically different.

Step 6: For your first regression of ECV on social group size, calculate the standard error for the slope coefficient, the 95% CI, and the p value associated with this coefficient by hand. Also extract this same information from the results of running the lm() function.
```{r}
n <- length(x)

y2 <- beta0 + beta1 * x
residuals <- y - y2

SSE <- sum(residuals^2)          # Sum of squared errors
residuals_var <- SSE / (n - 2)      # Residual variance
SE_beta1 <- sqrt(residuals_var / sum((x - x_mean)^2))
SE_beta1 #standard error of the slope

#95% CI
t_crit <- qt(0.975, df = n - 2)
CI_lower <- beta1 - t_crit * SE_beta1
CI_upper <- beta1 + t_crit * SE_beta1
CI_lower
CI_upper

#p-value
t_stat <- beta1 / SE_beta1
p_value <- 2 * (1 - pt(abs(t_stat), df = n - 2))

t_stat
p_value

#check using lm()
model <- lm(ECV ~ Group_size, data = d1)
summary(model)
```
Same results!

Step 7: Use a permutation approach with 1000 permutations to generate a null sampling distribution for the slope coefficient. You can use either the quantile method (i.e., using quantiles from the actual permutation-based null sampling distribution) or a theory-based method (i.e., using the standard deviation of the permutation-based null sampling distribution as the estimate of the standard error, along with a normal or t distribution), or both, to calculate this p value.
```{r}
set.seed(123)

d_perm <- d2 %>%
  drop_na(ECV, Group_size)

n_perm <- 1000
perm_slopes <- numeric(n_perm)

for (i in 1:n_perm) {
  
  y_perm <- sample(d_perm$ECV)  # permute response
  
  fit <- lm(y_perm ~ d_perm$Group_size)
  perm_slopes[i] <- coef(fit)[2]
}
obs_model <- lm(ECV ~ Group_size, data = d_perm)
obs_slope <- coef(obs_model)[2]
obs_slope

#quantile method
p_value_perm <- mean(abs(perm_slopes) >= abs(obs_slope))
p_value_perm

#standard deviation method
perm_se <- sd(perm_slopes)

z <- obs_slope / perm_se
p_value_norm <- 2 * (1 - pnorm(abs(z)))

p_value_norm

```
What is it that you need to permute? You need to permute the ECV values relative to Group_size (i.e., randomly shuffle ECV among observations while keeping Group_size fixed). This breaks any real relationship between brain size and social group size under the null hypothesis.

What is the p value associated with your original slope coefficient? Using the permutation test, the quantile-based p-value is 0 (no permuted slopes were as extreme as the observed slope). Using the standard deviation (normal approximation) method, the p-value is approximately 4.33 × 10⁻¹⁰.

Step 8:Use bootstrapping to generate a 95% CI for your estimate of the slope coefficient using both the quantile method and the theory-based method (i.e., using the standard deviation of the bootstrapped sampling distribution as an estimate of the standard error).
```{r}
set.seed(123)

d_boot <- d2 %>%
  drop_na(ECV, Group_size)

n <- nrow(d_boot)

boot_slopes <- numeric(1000)

for (i in 1:1000) {
  
  boot_data <- d_boot[sample(1:n, replace = TRUE), ]
  
  fit <- lm(ECV ~ Group_size, data = boot_data)
  
  boot_slopes[i] <- coef(fit)[2]
}

obs_model <- lm(ECV ~ Group_size, data = d_boot)
obs_slope <- coef(obs_model)[2]
obs_slope

CI_quantile <- quantile(boot_slopes, c(0.025, 0.975))
CI_quantile

boot_se <- sd(boot_slopes)

CI_lower <- obs_slope - 1.96 * boot_se
CI_upper <- obs_slope + 1.96 * boot_se

c(CI_lower, CI_upper)
```
Do these CIs suggest that your slope coefficient is different from zero?
Yes, both bootstrap confidence intervals (quantile: 1.41–3.27 and theory-based: 1.52–3.40) do not include 0, indicating that the slope is significantly different from zero. This suggests a significant positive relationship between ECV and social group size.